---
title: "GLM & Logistic Regression"
subtitle: "Generalized linear models"
description: "GLMs and logistic regression for non-normal responses."
categories: [GLM, Regression]
order: 14
week: 8
author: "Bill Perry"

execute: 
  collapse: true
knitr:
  opts_chunk:
    paged.print: false
    collapse: true
format:
  html:
    downloads: [docx, pptx]  # This creates download links for all three
  # RevealJS - UNCHANGED (keeps your two-column layout and large images)
  revealjs:
    output-ext: "slides.html"
  docx: default
  pptx: default
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

# Load required packages
# install.packages("ResourceSelection")
# install.packages("pscl")
library(pscl)
library(tinytable)
library(performance)
library(ResourceSelection)
library(car)
library(emmeans)
library(DHARMa)
library(MASS)
library(broom)
library(flextable)
library(patchwork)  # For combining plots
library(faraway)  # For gala dataset
library(tidyverse)


```

# Lecture 13: Review of ANOVAs

### Review

-   ANOVA
-   Factorial ANOVA
-   Nested ANOVA
-   ASSUMPTIONS OF ALL
    -   Homogeneity of variance - Levenes or Bartletts Test

    -   Normality of Residuals

    -   Independence

# Lecture 14: GLM Overview

### Overview

-   General Linear Models GLM
-   Essentially the same as before while using defined distributions
    -   Normal
    -   Lognormal
    -   Binomial
    -   Poisson
    -   Gamma
    -   Negative binomial
-   Logistic Regression
    -   when the outcome is yes or no or categorical

# Overview of Generalized Linear Models (GLMs)

::::: columns
::: {.column width="60%"}
General linear models assume normal distribution of response variables
and residuals. However, many types of biological data don't meet this
assumption.

Generalized Linear Models (GLMs) allow for a wider range of probability
distributions for the response variable.

GLMs allow all types of "exponential family" distributions:

-   Normal
-   Lognormal
-   Binomial
-   Poisson
-   Gamma
-   Negative binomial

GLMs can be used for binary (yes/no), discrete (count), and
categorical/multinomial response variables, using maximum likelihood
(ML) rather than ordinary least squares (OLS) for estimation.

**Note:** GLMs extend linear models to non-normal data distributions.
:::

::: {.column width="40%"}
```{r glm-distributions}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Examples of distributions in the exponential family
#| fig-height: 5
#| fig-width: 6
#| paged-print: false

# Create example data for different distributions
set.seed(123)
x <- seq(-4, 4, length.out = 100)
normal <- dnorm(x)
poisson_data <- dpois(0:10, lambda = 3)
binomial_data <- dbinom(0:10, size = 10, prob = 0.3)

# Plot the distributions
p1 <- ggplot(data.frame(x, normal), aes(x, normal)) +
  geom_line() +
  labs(title = "Normal Distribution",
       x = "Value", y = "Density") +
  theme_minimal()

p2 <- ggplot(data.frame(x = 0:10, y = poisson_data), 
             aes(x, y)) +
  geom_col(fill = "darkblue") +
  labs(title = "Poisson Distribution (λ = 3)",
       x = "Count", y = "Probability") +
  theme_minimal()

p3 <- ggplot(data.frame(x = 0:10, y = binomial_data), 
             aes(x, y)) +
  geom_col(fill = "darkred") +
  labs(title = "Binomial Distribution (n = 10, p = 0.3)",
       x = "Successes", y = "Probability") +
  theme_minimal()

(p1 / (p2 + p3))
```
:::
:::::

# Example comparison of logged response variable and GLM

::::: columns
::: {.column width="50%"}
```{r l14-01}
# Simulate data that students might log-transform
set.seed(123)
treatment_data <- data.frame(
  treatment = rep(c("Control", "Low", "High"), each = 20),
  biomass = c(rlnorm(20, meanlog = 2, sdlog = 0.5),
              rlnorm(20, meanlog = 2.5, sdlog = 0.5),
              rlnorm(20, meanlog = 3, sdlog = 0.5)))
# Approach 1: Log transformation + ANOVA
model_log <- lm(log(biomass) ~ treatment, data = treatment_data)

# Approach 2: Gamma GLM with log link
model_glm <- glm(biomass ~ treatment, 
                 family = Gamma(link = "log"), 
                 data = treatment_data)
```
:::

::: {.column width="50%"}
Here is the difference

```{r l14-02}
# Compare predictions on original scale
cat("logged response variable\n\n")
emmeans(model_log, ~treatment, type = "response")  # Geometric means
cat("\n\n")
cat("using glm opn original scale\n\n")
emmeans(model_glm, ~treatment, type = "response")  # Arithmetic means

```
:::
:::::

# The Three Elements of a GLM

### GLMs consist of three components:

1.  **Response component**: The response variable and its probability
    distribution (from exponential family: normal, binomial, Poisson)
2.  **Systematic component**: The predictor variable(s) in the model,
    which can be continuous or categorical
3.  **Link function**: Connects expected value of Y to predictor
    variables

$$g(\mu) = \beta_0 + \beta_1X_1 + \beta_2X_2...$$

::: {.callout-important appearance="simple"}
## Link Functions and Distributions

| Distribution | Common Link Function | Formula                      |
|--------------|----------------------|------------------------------|
| Normal       | Identity             | $g(\mu) = \mu$               |
| Poisson      | Log                  | $g(\mu) = \log(\mu)$         |
| Binomial     | Logit                | $g(\mu) = \log[\mu/(1-\mu)]$ |
:::

# GLM with Gaussian (Normal) Distribution: Setup

::::: columns
::: {.column width="60%"}
The simplest form of GLM uses a normal (Gaussian) distribution with an
identity link function. This is equivalent to standard ANOVA

-   Let's compare a standard linear model and a Gaussian GLM

-   Island Biogeography Data

-   The `gala` dataset from the `faraway` package contains data on 30
    Galapagos islands, testing MacArthur-Wilson's theory of island
    biogeography.

-   **Variables in the dataset:**

    -   `Species`- Number of plant species (count data)

    -   `Endemics`- Number of endemic species (count data)

    -   `Area`- Island area (km²)

    -   `Elevation`- Maximum elevation (m)

    ```{r gaussian-model}
    #| echo: false
    #| message: false
    #| warning: false
    #| paged-print: false

    # Create a categorical variable for demonstration
    gala <- gala %>%
      mutate(size_category = case_when(
        Area < 1 ~ "Small",
        Area >= 1 & Area < 100 ~ "Medium",
        Area >= 100 ~ "Large"
      ),
      size_category = factor(size_category, levels = c("Small", "Medium", "Large")))

    head(gala %>% dplyr::select(-Scruz, -Adjacent),3)
    ```
:::

::: {.column width="40%"}
Let's look at the summary of our Gaussian GLM:

```{r summary-gaussian_2}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

ggplot(gala, aes(x = Endemics)) +
  geom_histogram(binwidth = 5, fill = "darkblue", color = "black") +
  labs(title = "Distribution of Endemics Richness",
       subtitle = "Galapagos Islands",
       x = "Number of Plant Endemics",
       y = "Number of Islands") +
  theme_minimal()
```
:::
:::::

# GLM with Gaussian (Normal) Distribution: Setup

::::: columns
::: {.column width="60%"}
The simplest form of GLM uses a normal (Gaussian) distribution with an
identity link function. This is equivalent to standard linear
regression.

Let's compare a standard linear model and a Gaussian GLM using the
Galapagos dataset, modeling endemic Endemics richness by island size
category.

```{r gaussianmodel2a}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

gala %>% arrange(-Area) %>% dplyr::select(Endemics, Area, size_category, Elevation) %>% head(8) %>%
  tt()
```
:::

::: {.column width="40%"}
Let's visualize endemic Endemics by island size:

```{r summarygaussian_2a}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

gala %>% ggplot(aes(size_category, Endemics, fill = size_category)) + 
  geom_boxplot() + 
  theme(
    axis.text = element_text(size=14),
    axis.title = element_text(size=14)
  ) +
  labs(x = "Island Size Category", 
       y = "Number of Endemic Endemics",
       color = "Size Category")
```
:::
:::::

# GLM with Gaussian (Normal) Distribution: Setup

::::: columns
::: {.column .scrollable width="50%"}
The simplest form of GLM uses a normal (Gaussian) distribution with an
identity link function. This is equivalent to standard linear
regression.

-   Let's compare a standard linear model and a Gaussian GLM using the
    Galapagos dataset, modeling endemic species richness by island size
    category.
    -   Fit a standard linear model - model_lm \<- lm(Endemics \~
        size_category,\
        data = gala)
    -   Fit a Gaussian GLM
    -   model_gaussian \<- glm(Endemics \~ size_category, data = gala,\
        family = gaussian(link = "identity")) **Residual Standard
        Error**. This is the standard deviation of the residuals
        ($\sigma$)
:::

::: {.column .scrollable width="50%"}
-   Let's look at the summary of our Gaussian GLM:

```{r l14-03}
#| include: false
# Fit a standard linear model
model_lm <- lm(Endemics ~ size_category, data = gala)
# Fit a Gaussian GLM
model_gaussian <- glm(Endemics ~ size_category,  data = gala, 
                       family = gaussian(link = "identity"))
# # Compare the coefficients
coef_lm <- coefficients(model_lm)
coef_glm <- coefficients(model_gaussian)
# # Check if they're the same
# all.equal(coef_lm, coef_glm)
```

```{r gaussian-model_2a}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
summary(model_lm)
```
:::
:::::

# GLM with Gaussian (Normal) Distribution: Setup

::::: columns
::: {.column width="60%"}
-   Let's look at the summary of our Gaussian GLM:
    -   Dispersion parameter
        -   **variance of the residuals** ($\sigma^2$)
    -   Null deviance
        -   This is the error (deviance) of the "Null" or
            "Intercept-Only" model

        -   how much error you have when you try to predict the number
            of endemics using *only* the overall average, without any
            predictors. It's your baseline, or the total amount of
            variation in the data to be explained
    -   Residual deviance
        -   conceptually the same as the "Model Sum of Squares" in an
            ANOVA

        -   Null Deviance tells you your `size_category` predictor is
            *very* useful for explaining the number of endemics.
:::

::: {.column width="40%"}
```{r summary-gaussian3b}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
summary(model_gaussian)
```
:::
:::::

# GLM with Gaussian Distribution: Analysis

::::: columns
::: {.column width="60%"}
-   Now let's perform an ANOVA on LM and GLM model using `car`:
    -   **Residual Standard Error**. This is the standard deviation of
        the residuals ($\sigma$)

    -   **Residual Deviance/df**: 5756.1 / 27 = 213.1878

    -   In this context, "Sum Sq" (Sum of Squares) and "Deviance" are
        the same thing.

    -   **"Pearson residuals" are essentially the same as the regular
        residuals**

```{r gaussian-anovaa}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
Anova(model_lm, type = "III", test = "F")
Anova(model_gaussian, type = "III", test = "F")
```
:::

::: {.column width="40%"}
Visualizing the results:

```{r gaussian-plot}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false


# Get estimated means
emm_gaussian <- emmeans(model_gaussian, ~ size_category)
emm_df <- as.data.frame(emm_gaussian)

# Create plot of data with estimated means
ggplot() +
  # Plot raw data
  geom_jitter(data = gala, 
              aes(x = size_category, y = Endemics), 
              width = 0.2, 
              alpha = 0.5) +
  # Add estimated means with confidence intervals
  geom_point(data = emm_df, 
             aes(x = size_category, y = emmean), 
             size = 4, color = "red") +
  geom_errorbar(data = emm_df, 
                aes(x = size_category, 
                    ymin = lower.CL, 
                    ymax = upper.CL), 
                width = 0.2, 
                color = "red") +
  labs(
       x = "Size",
       y = "# Endemics") +
  theme_minimal() + theme(
  axis.text = element_text(size=14) ,
  axis.title = element_text(size=14) ) 
```
:::
:::::

# Equivalence of Linear Models and Gaussian GLMs

### Equivalence of Linear Models and Gaussian GLMs

When we use a **Gaussian distribution** with an **identity link**, GLM
gives identical results to standard linear regression. This can be seen
in the coefficient values and overall model statistics.

The key difference is that GLMs provide a framework that extends to
non-normal distributions.

[**Note when we look at Heteroscedascidity (SP) cone of despair - this
is why a Poisson model works**]{.underline}

# GLM with Poisson Distribution: Setup

-   **Poisson GLMs** Poisson model used when response variable is
    **count data**:
    -   Number of species on an island
    -   Number of parasites in a host
    -   Number of bird nests in a plot
    -   Number of seeds produced by a plant
-   The Poisson distribution assumes:
    -   Counts are non-negative integers (0, 1, 2, 3, ...)
    -   The mean equals the variance
    -   Events occur independently
-   **Key consideration:** If variance \> mean (overdispersion),
    consider negative binomial regression instead.
-   Now let's fit a Poisson GLM to model the relationship between the
    rounded quarter-mile time and the number of cylinders:

# Fit Poisson GLM with size_category as predictor

::::: columns
::: {.column width="50%"}
-   model is predicting the natural logarithm (log) of the expected
    species count
-   exponentiate them (i.e., calculate $e^{\text{coefficient}}$ or
    exp(coefficient).
-   converts the additive log-counts back into a multiplicative effect
    on the actual species count
-   exp(2.67101) = 14.45 species on a "Small" island
-   exp(1.33784) = 3.81
    -   "Medium" predicted to have 3.81x more spp than "Small" island
-   exp(2.84300) = 17.17
    -   "Large" island is predicted to have 17.17 times more species
        than a "Small" island, on average
-   Dispersion - Divide the deviance by its df: $939.74 / 27 \approx$
    34.8
:::

::: {.column width="50%"}
```{r l14-04}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
model_poisson_gala <- glm(Endemics ~ size_category, 
                          data = gala,
                          family = poisson(link = "log"))
summary(model_poisson_gala)
```
:::
:::::

# GLM with Poisson Distribution: Setup

::::: columns
::: {.column width="50%"}
Does island size category, as a whole, have a statistically significant
effect on the number of plant species?

-   `test = "LR"`: important part!
    -   normal ANOVA (with a Gaussian/normal distribution) test is an
        F-test.
    -   GLM (like Poisson) can't use F-test in the same way
        -   use a Likelihood Ratio (LR) test
        -   LR test statistically compares fit of full model (the one
            with size_category) to simpler null model (one without
            size_category)
        -   LR test tells us if it is significant
:::

::: {.column width="50%"}
Test overall effect of size_category

```{r l14-05}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false

Anova(model_poisson_gala, type = "III", test = "LR")
```
:::
:::::

# GLM with Poisson Distribution: Setup

::::: columns
::: {.column width="50%"}
**Poisson GLMs** appropriate for **count data** and assumes variance
equals mean - use the Gala Data and endemics - With natural log link,
coefficients = **multiplicative effects**: - A coefficient of β means:
for each 1-unit increase in X, the response is multiplied by exp(β) -
For small β, exp(β) ≈ 1 + β, so β × 100% =app. %change

```{r poisson-model}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
summary(model_poisson_gala)
```
:::

::: {.column width="50%"}
-   check overdispersion- common in count data:
-   Should be close to 1 for well-fitting Poisson model
-   If \> 1.5, may indicate overdispersion
-   Dispersion parameter measures obs var / expected var:
-   **Dispersion ≈ 1**: Good fit (var = mean)
-   **Dispersion \> 1**: Overdispersion (var \> mean)
-   **Dispersion \< 1**: **Underdispersion** (var \< mean)
-   use Negative Binomial model

```{r gala-poisson-dispersion}
#| message: false
#| warning: false
# Calculate the dispersion parameter
# (Pearson's Chi-Squared statistic / residual degrees of freedom)
dispersion_gala <- sum(residuals(model_poisson_gala, type = "pearson")^2) / 
                   model_poisson_gala$df.residual

# Print dispersion parameter
cat("Dispersion parameter:", round(dispersion_gala, 2), "\n")
```
:::
:::::

# GLM with Poisson Distribution: Setup

::::: columns
::: {.column width="60%"}
-   **Poisson GLMs** are appropriate for **count data**
-   assumes that the variance equals the mean.

```{r gala-poisson-emmeans}
#| message: false
#| warning: false

# 1. Calculate Estimated Marginal Means (EMMs)
# type = "response" converts the log-means back to the "Endemics count" scale
emm_gala <- emmeans(model_poisson_gala, 
                    specs = ~ size_category,
                    type = "response")
print(emm_gala)
```
:::

::: {.column width="40%"}
Let's check the emmeans and pairwise comparisons

```{r poisson-overdispersion}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# 2. Pairwise Comparisons
# Compares all pairs: Small-Medium, Small-Large, Medium-Large
pairs_gala <- pairs(emm_gala, adjust = "tukey")
print(pairs_gala)

# 3. Compact Letter Display (CLD)
# The easiest way to see the groupings
cld_gala <- multcomp::cld(emm_gala, 
                          Letters = letters,  
                          alpha = 0.05)
print(cld_gala)

```
:::
:::::

# Poisson GLM: Visualization and Interpretation

::::: columns
::: {.column width="60%"}
-   Plot raw data and our emmeans results

    -   shows the model's predictions on top of the real data

    ```{r poisson-plot}
    #| echo: false
    #| message: false
    #| warning: false
    #| paged-print: false


    # 1. Get the estimated means and CIs into a dataframe
    emm_gala_df <- as.data.frame(emm_gala)

    # 2. Create visualization
    ggplot() +
      # Plot raw data (jittered so we can see the points)
      geom_jitter(data = gala,
                  aes(x = size_category, y = Endemics),
                  width = 0.2, # Spreads points horizontally
                  alpha = 0.5) +
      # Add estimated means (points)
      geom_point(data = emm_gala_df, 
                 aes(x = size_category, y = rate), # 'rate' is the mean
                 size = 4, color = "blue") +
      # Add confidence intervals (error bars)
      geom_errorbar(data = emm_gala_df, 
                    aes(x = size_category, 
                        ymin = asymp.LCL, # Lower Confidence Limit
                        ymax = asymp.UCL), # Upper Confidence Limit
                    width = 0.2, color = "blue", linewidth = 1) +
      labs(title = "Endemics Richness by Island Size Category",
           subtitle = "Poisson GLM predictions (on the response scale)",
           x = "Island Size Category",
           y = "Number of Plant Endemics") +
      theme_minimal()
    ```
:::

::: {.column width="40%"}
### Interpreting Poisson GLM Coefficients

-   In a Poisson GLM with a log link function:
    -   Coefficientschanges in the **log** of the expected count
    -   When exponentiated (`exp(coef)`)= multiplicative effects
    -   `exp(coef)` = 0.90 = the expected count is 90% of reference
:::
:::::

# Comparison of interpretation

-   Interpretation of β₁ GLM (Log-Link) glm(Y\~X, family = poisson)
    -   1-unit change in X multiplies the mean of Y by exp(β₁)
-   lm (Log-Response) lm(log(Y) \~ X)
    -   1-unit change in X multiplies median Y by exp(β₁).
-   lm (Log-Log) lm(log(Y) \~ log(X))
    -   A 1% change in x is a beta percent change in y...

| Model Type | R Code (Example) | Interpretation of β₁ |
|----|----|----|
| GLM (Log-Link) | `glm(Y ~ X, family = poisson)` | A 1-unit change in X multiplies the mean of Y by exp(β₁). |
| lm (Log-Response) | `lm(log(Y) ~ X)` | A 1-unit change in X multiplies the median of Y by exp(β₁). |
| lm (Log-Log) | `lm(log(Y) ~ log(X))` | A 1% change in X is associated with a β₁% change in Y. |

# Checking Model Assumptions with DHARMa

::: {.panel column="screen"}
```{r gala-poisson-diagnostics}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 6

# Load DHARMa library (if not already loaded)
# library(DHARMa) 

# Simulate residuals
set.seed(123) # For reproducible results
simulation_gala <- simulateResiduals(fittedModel = model_poisson_gala, n = 1000)

# Plot diagnostic plots
plot(simulation_gala)
```
:::

# Dealing with Overdispersion in Count Data

::::: columns
::: {.column width="60%"}
When overdispersion (variance \> mean) - use negative binomial model
`model_nb <- glm.nb(Endemics ~ area_cat, data = gala)`

-   negative binomial model includes a dispersion parameter (theta)

    -   allows variance to be larger than mean
    -   **standard errors bigger because NB model accounts for high
        variability (overdispersion)**
    -   estimates dispersion parameter 'Theta' (or 1/theta)

    ```{r negbin-model}
    #| echo: false
    #| message: false
    #| warning: false
    #| paged-print: false
    model_nb_gala <- glm.nb(Endemics ~ size_category, data = gala)
    summary(model_nb_gala)
    ```
:::

::: {.column width="40%"}
Let's compare the predictions from both models:

```{r gala-compare-models}
#| echo: false
#| message: false
#| warning: false

# Add predictions from both models to the gala dataframe
gala$pred_poisson <- predict(model_poisson_gala, type = "response")
gala$pred_nb <- predict(model_nb_gala, type = "response")

# Plot Poisson predictions vs. Negative Binomial predictions
ggplot(gala, aes(x = pred_poisson, y = pred_nb, color = size_category)) +
  geom_point(size = 3, alpha = 0.7) +
  # Add a 1-to-1 reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(title = "Comparison of Poisson and Negative Binomial Predictions",
       x = "Poisson Predicted Endemics Count",
       y = "Negative Binomial Predicted Endemics Count",
       color = "Size Category") +
  theme_minimal()
```
:::
:::::

# Logistic Regression - Introduction

::::: columns
::: {.column width="60%"}
Logistic regression is a GLM used when the response variable is binary
(e.g., dead/alive, present/absent). It models the probability of the
response being "1" (success) given predictor values.

-   Logistic regression model:
    -   $\pi(x) = \frac{e^{\beta_0 + \beta_1 x}}{1 + e^{\beta_0 + \beta_1 x}}$
    -   Where:
    -   $\pi(x)$ is the probability that Y = 1 given X = x
    -   $\beta_0$ is the intercept
    -   $\beta_1$ is the slope (rate of change in $\pi(x)$ for a unit
        change in X)
-   To linearize this relationship, we use the logit link function:
    -   $g(x) = \log\left(\frac{\pi(x)}{1-\pi(x)}\right) = \beta_0 + \beta_1 x$
-   transforms probability (bounded between 0 and 1) to linear fx ranges
    from -∞ to +∞.
:::

::: {.column width="40%"}
```{r logistic-curve}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# Create data for sigmoid curve
sigmoid_data <- data.frame(
  x = seq(-6, 6, length.out = 100)
)
sigmoid_data$p <- 1 / (1 + exp(-sigmoid_data$x))

# Plot the sigmoid curve
ggplot(sigmoid_data, aes(x, p)) +
  geom_line(linewidth = 1.2, color = "darkblue") +
  geom_hline(yintercept = c(0, 0.5, 1), 
             linetype = "dashed", 
             color = "gray50") +
  geom_vline(xintercept = 0, 
             linetype = "dashed", 
             color = "gray50") +
  labs(title = "Logistic Function",
       subtitle = "Mapping from linear predictor to probability",
       x = "Linear predictor (β0 + β1x)",
       y = "Probability π(x)") +
  scale_y_continuous(breaks = seq(0, 1, 0.25)) +
  theme_minimal()
```
:::
:::::

# Example: Lizard Presence on Islands

Based on the example from Polis et al. (1998) - presence/absence of
lizards (*Uta*) on islands in Gulf of California based on perimeter/area
ratio.

::: {.panel column="screen"}
```{r lizard-data}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
set.seed(123)
island_data <- data.frame(
  island_id = 1:30,
  pa_ratio = seq(5, 70, length.out = 30),
  uta_present = c(rep(1, 10), 
                  rbinom(10, 1, prob = 0.5),  # Mixed outcomes in middle
                  rep(0, 10)))%>%
  mutate(uta_present_factor = factor(uta_present, levels = c(0, 1), 
         labels = c("Absent", "Present")))
ggplot() +
  # Add jittered points for observed data
  geom_point(data = island_data, 
              aes(x = pa_ratio, y = uta_present),
              position = position_dodge2(width=.1), alpha = 0.7) +
  labs(title = "Probability of Uta Presence vs. Perimeter/Area Ratio",
       x = "Perimeter/Area Ratio",
       y = "Probability of Presence") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()

```
:::

# Example: Lizard Presence on Islands

Based on the example from Polis et al. (1998) - presence/absence of
lizards (*Uta*) on islands in Gulf of California based on perimeter/area
ratio.

::: {.panel column="screen"}
```{r l14-06}
# Fit the logistic regression model
lizard_model <- glm(uta_present ~ pa_ratio, 
                    data = island_data, 
                    family = binomial(link = "logit"))

# Model summary
summary(lizard_model)
```
:::

# Lizard Example: Visualization and Testing

::::: columns
::: {.column width="60%"}
Let's visualize the data and the fitted model:

```{r lizard-plot}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# Create a dataframe for predictions
pred_data <- data.frame(
  pa_ratio = seq(min(island_data$pa_ratio), 
                max(island_data$pa_ratio), 
                length.out = 100)
)

# Get predicted probabilities
pred_data$prob <- predict(lizard_model, 
                         newdata = pred_data, 
                         type = "response")

# Plot
ggplot() +
  # Add jittered points for observed data
  geom_point(data = island_data, 
              aes(x = pa_ratio, y = uta_present),
              position = position_dodge2(width=.1), alpha = 0.7) +
  # Add predicted probability curve
  geom_line(data = pred_data, 
            aes(x = pa_ratio, y = prob), 
            color = "blue", size = 1) +
  # Add confidence intervals (optional)
  labs(title = "Probability of Uta Presence vs. Perimeter/Area Ratio",
       x = "Perimeter/Area Ratio",
       y = "Probability of Presence") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()
```
:::

::: {.column width="40%"}
-   Test null hypothesis that β₁ = 0
    -   no relationship between P/A ratio and lizard presence
-   There are two common ways to test this hypothesis:

1.  **Wald test**: Tests if the parameter estimate divided by its
    standard error differs significantly from zero
2.  **Likelihood ratio test**: Compares the fit of the full model to a
    reduced model without the predictor variable

```{r likelihood-ratio-test}
reduced_model <- glm(uta_present ~ 1, 
                     data = island_data, 
                     family = binomial(link = "logit"))
anova(reduced_model, lizard_model, test = "Chisq")
```
:::
:::::

# Interpreting the Odds Ratio

### Working with Odds Ratios

Odds ratio represents how odds of event (e.g., lizard presence) change
with unit increase in predictor

-   **Odds ratio = exp(β₁)**
-   \> 1: Increasing predictor increases odds of event
-   \< 1: Increasing predictor decreases odds of event
-   = 1: No effect of predictor on odds of event
-   one-unit increase in island's Perimeter/Area Ratio - odds of finding
    lizard present multiplied by 0.898
-   odds decrease by 10.2% (1-0.898) for every one-unit increase P/A
-   entire interval is below 1.0, relationship is negative: more P/A
    ratio means lower odds of lizards

::: {.panel column="screen"}
```{r odds-ratio}
coef_lizard <- coef(lizard_model)[2]  # Extract slope coefficient
odds_ratio <- exp(coef_lizard)
ci <- exp(confint(lizard_model, "pa_ratio"))
cat("Odds Ratio:", round(odds_ratio, 3), "\n\n",
"95% CI:", round(ci[1], 3), "to", round(ci[2], 3), "\n")
```
:::

# Assessing Model Fit

There are several ways to assess the goodness-of-fit for logistic
regression models:

::: {.panel column="screen"}
```{r model-fit}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# Calculate Hosmer-Lemeshow statistic
# This would normally require an additional package like 'ResourceSelection'
# Instead, we'll use a simpler approximation and other diagnostics

# Calculate Pearson residuals
pearson_resid <- residuals(lizard_model, type = "pearson")
pearson_chi2 <- sum(pearson_resid^2)
df_resid <- lizard_model$df.residual
# Calculate deviance
deviance_g2 <- lizard_model$deviance
null_deviance <- lizard_model$null.deviance
# Calculate McFadden's pseudo-R²
r2_mcfadden <- 1 - (deviance_g2 / null_deviance)
# Display results

cat("\n\nPearson χ²:", round(pearson_chi2, 3), "on", df_resid, "df, p =", 
    round(1 - pchisq(pearson_chi2, df_resid), 3), "\n\n")
cat("\n\n")
cat("\n\nDeviance G²:", round(deviance_g2, 3), "on", df_resid, "df, p =", 
    round(1 - pchisq(deviance_g2, df_resid), 3), "\n\n")
cat("\n\n")
cat("\n\nMcFadden's R²:", round(r2_mcfadden, 3), "\n\n")
# This one function gives McFadden's and other popular R² values
performance::r2(lizard_model)
cat("\n\n")
pscl::pR2(lizard_model)
# or 
cat("\n\n")
cat("\n\n")
# glance() gives a one-row data frame of all model-level stats
glance(lizard_model)
cat("\n\n")
# 'g' is the number of groups to test (e.g., 10 for deciles)
# Note: You need to provide the observed 'y' values
hoslem.test(lizard_model$y, fitted(lizard_model), g = 10)
cat("\n\n")
```
:::

# Multiple Logistic Regression: Setup

::::: columns
::: {.column width="50%"}
Logistic regression can be extended to include multiple predictors. The
model becomes:

$$g(x) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_p x_p$$

Where g(x) is the logit link function, and x₁, x₂, ..., xₚ are the
predictor variables.

Let's create a simulated dataset based on the Bolger et al. (1997) study
of the presence/absence of native rodents in canyon fragments.

```{r fragment-data}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| 
# Simulate data for the rodent example
set.seed(123)
n <- 25  # 25 canyon fragments

# Create predictor variables
fragment_data <- data.frame(
  fragment_id = paste0("F", 1:n),
  distance = runif(n, 0, 3000),            # Distance to source canyon (m)
  age = runif(n, 5, 80),                   # Years since isolation
  shrub_cover = runif(n, 10, 100)          # Percentage shrub cover
)

# Generate response variable (rodent presence)
# Higher probability with higher shrub cover, slight effect of age
linear_pred <- -5 + 0.0001*fragment_data$distance + 
               0.02*fragment_data$age + 
               0.09*fragment_data$shrub_cover
prob <- 1 / (1 + exp(-linear_pred))
fragment_data$rodent_present <- rbinom(n, 1, prob)
fragment_data$rodent_present <- factor(fragment_data$rodent_present, 
                                      levels = c(0, 1), 
                                      labels = c("Absent", "Present"))

# Fit multiple logistic regression model
rodent_model <- glm(rodent_present ~ distance + age + shrub_cover, 
                    data = fragment_data, 
                    family = binomial(link = "logit"))

# Model summary
summary(rodent_model)
```
:::

::: {.column width="50%"}
To test the significance of individual predictors, we can use likelihood
ratio tests comparing nested models:

```{r test-predictors}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# Test distance
model_no_distance <- glm(rodent_present ~ age + shrub_cover, 
                         data = fragment_data, 
                         family = binomial(link = "logit"))
anova(model_no_distance, rodent_model, test = "Chisq")

# Test age
model_no_age <- glm(rodent_present ~ distance + shrub_cover, 
                    data = fragment_data, 
                    family = binomial(link = "logit"))
anova(model_no_age, rodent_model, test = "Chisq")

# Test shrub cover
model_no_shrub <- glm(rodent_present ~ distance + age, 
                      data = fragment_data, 
                      family = binomial(link = "logit"))
anova(model_no_shrub, rodent_model, test = "Chisq")
```
:::
:::::

# Multiple Logistic Regression: Odds Ratios

Let's calculate odds ratios and confidence intervals for all predictors.

::: {.panel column="screen"}
```{r multiple-odds-ratios}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# Calculate odds ratios and CIs
coefs <- coef(rodent_model)[-1]  # Exclude intercept
odds_ratios <- exp(coefs)
ci <- exp(confint(rodent_model)[-1, ])  # Exclude intercept

# Create a data frame for display
or_df <- data.frame(
  Predictor = names(coefs),
  OddsRatio = odds_ratios,
  LowerCI = ci[, 1],
  UpperCI = ci[, 2]
)

# Display formatted table
or_df %>%
  mutate(across(where(is.numeric), round, 4)) %>%
  mutate(CI = paste0("(", LowerCI, ", ", UpperCI, ")")) %>%
  dplyr::select(Predictor, OddsRatio, CI) 
```
:::

# Visualizing Multiple Logistic Regression

::: {.panel column="screen"}
For multiple predictors, we can visualize the effect of each predictor
while holding others constant at their mean or median values.

This visualization shows the effect of each predictor on the probability
of rodent presence, while holding the other predictors constant at their
mean values.

```{r visualize-effects}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 4
#| fig-width: 8
#| paged-print: false

# Create a function to generate prediction data for one variable
predict_for_var <- function(var_name, model, data) {
  # Create grid of values for the variable of interest
  pred_df <- data.frame(
    x = seq(min(data[[var_name]]), max(data[[var_name]]), length.out = 100)
  )
  names(pred_df) <- var_name
  
  # Add mean values for other predictors
  for (other_var in c("distance", "age", "shrub_cover")) {
    if (other_var != var_name) {
      pred_df[[other_var]] <- mean(data[[other_var]])
    }
  }
  
  # Add predictions
  pred_df$prob <- predict(model, newdata = pred_df, type = "response")
  
  return(pred_df)
}

# Generate prediction data for each variable
pred_distance <- predict_for_var("distance", rodent_model, fragment_data)
pred_age <- predict_for_var("age", rodent_model, fragment_data)
pred_shrub <- predict_for_var("shrub_cover", rodent_model, fragment_data)

# Create plots
p1 <- ggplot() +
  geom_rug(data = fragment_data, 
           aes(x = distance, y = as.numeric(rodent_present) - 1),
           sides = "b", alpha = 0.7) +
  geom_line(data = pred_distance, aes(x = distance, y = prob), 
            color = "darkred", size = 1) +
  labs(title = "Effect of Distance",
       x = "Distance to Source (m)",
       y = "Probability of Presence") +
  theme_minimal()

p2 <- ggplot() +
  geom_rug(data = fragment_data, 
           aes(x = age, y = as.numeric(rodent_present) - 1),
           sides = "b", alpha = 0.7) +
  geom_line(data = pred_age, aes(x = age, y = prob), 
            color = "darkgreen", size = 1) +
  labs(title = "Effect of Age",
       x = "Years Since Isolation",
       y = "Probability of Presence") +
  theme_minimal()

p3 <- ggplot() +
  geom_rug(data = fragment_data, 
           aes(x = shrub_cover, y = as.numeric(rodent_present) - 1),
           sides = "b", alpha = 0.7) +
  geom_line(data = pred_shrub, aes(x = shrub_cover, y = prob), 
            color = "darkblue", size = 1) +
  labs(title = "Effect of Shrub Cover",
       x = "Shrub Cover (%)",
       y = "Probability of Presence") +
  theme_minimal()

# Combine plots
p1 + p2 + p3
```
:::

# Assumptions and Diagnostics of Logistic Regression

::::: columns
::: {.column width="50%"}
Logistic regression has several key assumptions:

1.  Independence of observations
2.  Linear relationship between predictors and log odds
3.  No extreme outliers
4.  No multicollinearity (when multiple predictors are used)

diagnostics for our multiple logistic regression model:
:::

::: {.column width="50%"}
```{r diagnostics}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 6
#| paged-print: false

# 1. Check for linearity between predictors and log odds
# Use bins of X variables and plot log odds
check_linearity <- function(model, data, var) {
  # Create bins of predictor
  n_bins <- 5
  data$bin <- cut(data[[var]], breaks = n_bins)
  
  # Calculate log odds for each bin
  bin_summary <- data %>%
    group_by(bin) %>%
    summarize(
      n = n(),
      mean_var = mean(!!sym(var)),
      successes = sum(rodent_present == "Present"),
      failures = sum(rodent_present == "Absent")
    ) %>%
    mutate(
      p = successes / n,
      logodds = log(p / (1 - p))
    )
  
  # Create plot
  ggplot(bin_summary, aes(x = mean_var, y = logodds)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste("Linearity Check:", var),
         x = var,
         y = "Log Odds") +
    theme_minimal()
}

# Create diagnostic plots for each variable
p1 <- check_linearity(rodent_model, fragment_data, "distance")
p2 <- check_linearity(rodent_model, fragment_data, "age")
p3 <- check_linearity(rodent_model, fragment_data, "shrub_cover")

# Arrange the plots
p1 / p2 / p3
```
:::
:::::

# Model Comparison and Selection

::::: columns
::: {.column width="60%"}
When working with multiple predictors, want to find most parsimonious
model. use: - Likelihood ratio tests for nested models - Information
criteria (AIC, BIC) for non-nested models - Classification metrics like
accuracy, sensitivity, and specificity

Compare models and AIC values:

```{r model-selection}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# Calculate AIC for our models
models <- list(
  "Full" = rodent_model,
  "No Distance" = model_no_distance,
  "No Age" = model_no_age,
  "No Shrub" = model_no_shrub,
  "Intercept Only" = glm(rodent_present ~ 1, 
                        data = fragment_data, 
                        family = binomial)
)

# Calculate AIC and BIC
model_comparison <- data.frame(
  Model = names(models),
  Parameters = sapply(models, function(m) length(coef(m))),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  Deviance = sapply(models, function(m) m$deviance)
)

# Show model comparison table
model_comparison %>%
  arrange(AIC) %>%
  mutate(across(where(is.numeric), round, 2)) 
```
:::

::: {.column width="40%"}
Evaluate predictive performance of our model:

```{r prediction-performance}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# Get predictions
predicted_probs <- predict(rodent_model, type = "response")
predicted_class <- ifelse(predicted_probs > 0.5, "Present", "Absent")

# Create confusion matrix
true_class <- fragment_data$rodent_present
conf_matrix <- table(Predicted = predicted_class, Actual = true_class)

# Calculate metrics
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
sensitivity <- conf_matrix["Present", "Present"] / sum(conf_matrix[, "Present"])
specificity <- conf_matrix["Absent", "Absent"] / sum(conf_matrix[, "Absent"])

# Display results
conf_matrix
cat("\nAccuracy:", round(accuracy, 3), "\n")
cat("Sensitivity:", round(sensitivity, 3), "\n")
cat("Specificity:", round(specificity, 3), "\n")
```
:::
:::::

# Publication-Quality Figure

Let's create a publication-quality figure for our multiple logistic
regression model and show how we would write up the results for a
scientific publication.

::: {.panel column="screen"}
```{r publication-figure}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false

# Create a more polished visualization for shrub cover effect
polished_pred <- predict_for_var("shrub_cover", rodent_model, fragment_data)

# Calculate confidence intervals
pred_se <- predict(rodent_model, 
                  newdata = polished_pred, 
                  type = "link", 
                  se.fit = TRUE)

# Convert to data frame with CIs
ci_data <- data.frame(
  shrub_cover = polished_pred$shrub_cover,
  fit = pred_se$fit,
  se = pred_se$se.fit
)

# Calculate upper and lower bounds of CI on link scale
ci_data$lower_link <- ci_data$fit - 1.96 * ci_data$se
ci_data$upper_link <- ci_data$fit + 1.96 * ci_data$se

# Transform back to probability scale
ci_data$prob <- plogis(ci_data$fit)
ci_data$lower_prob <- plogis(ci_data$lower_link)
ci_data$upper_prob <- plogis(ci_data$upper_link)

# Create plot
ggplot() +
  # Add jittered points for raw data
  geom_jitter(data = fragment_data, 
             aes(x = shrub_cover, 
                 y = as.numeric(rodent_present == "Present")),
             width = 0, height = 0.05, alpha = 0.6, size = 3) +
  # Add fitted probability curve
  geom_line(data = ci_data, 
           aes(x = shrub_cover, y = prob), 
           color = "darkblue", size = 1.2) +
  # Add confidence intervals
  geom_ribbon(data = ci_data, 
             aes(x = shrub_cover, 
                 ymin = lower_prob, 
                 ymax = upper_prob), 
             alpha = 0.2, fill = "darkblue") +
  # Customize appearance
  labs(title = "Effect of Shrub Cover on Native Rodent Presence",
       subtitle = "Probability of occurrence in canyon fragments",
       x = "Percentage Shrub Cover",
       y = "Probability of Rodent Presence") +
  scale_y_continuous(limits = c(0, 1), 
                     breaks = seq(0, 1, 0.2)) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(fill = NA, color = "gray80")
  )
```
:::

# Scientific Write-Up Example

::: {.callout-important appearance="simple"}
## Scientific Write-Up Example

**Results**

The presence of native rodents in canyon fragments was modeled using
multiple logistic regression with three predictors: distance to nearest
source canyon, years since isolation, and percentage of shrub cover. The
model was statistically significant (χ² = 12.63, df = 3, p = 0.005) and
explained 38.7% of the variation in rodent presence (McFadden's R² =
0.387).

Among the predictors, only shrub cover had a statistically significant
effect on rodent presence (β = 0.091, SE = 0.041, p = 0.026). The odds
ratio for shrub cover was 1.095 (95% CI: 1.011-1.186), indicating that
for each percentage increase in shrub cover, the odds of rodent presence
increased by approximately 9.5%. Neither distance to source canyon (β =
0.0002, p = 0.690) nor years since isolation (β = 0.022, p = 0.566)
showed significant relationships with rodent presence.

The model correctly classified 76% of the fragments, with a sensitivity
of 0.77 and a specificity of 0.75. Diagnostics indicated no significant
issues with model fit (Hosmer-Lemeshow χ² = 7.31, df = 8, p = 0.504).
:::

# Scientific Write-Up Example

::: {.callout-important appearance="simple"}
## Scientific Write-Up Example

**Discussion**

Our findings suggest that vegetation structure, as measured by shrub
cover, plays a crucial role in determining the presence of native
rodents in canyon fragments. The positive relationship between shrub
cover and rodent occurrence likely reflects the importance of vegetation
for providing food resources, shelter from predators, and suitable
microhabitat conditions. Contrary to our expectations, isolation metrics
(distance to source canyon and years since isolation) did not
significantly predict rodent presence, suggesting that local habitat
quality may be more important than landscape connectivity for these
species.
:::

# Relationship Between GLMs and ANOVAs

### GLMs and ANOVAs: The Connection

General linear models (including ANOVAs and standard regression) are
special cases of Generalized Linear Models where:

1.  The response variable follows a normal distribution
2.  The link function is the identity function

-   Therefore, a one-way ANOVA is equivalent to:
    -   A linear regression with a categorical predictor

    -   A Gaussian GLM with an identity link and a categorical predictor

# Demonstrating ANOVA-GLM Equivalence

Let's demonstrate this equivalence:

::: {.panel column="screen"}
```{r anova-glm-comparison}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 4
#| fig-width: 7
#| paged-print: false

# 1. Standard ANOVA
anova_model <- aov(mpg ~ cyl, data = mtcars)

# 2. Linear regression
lm_model <- lm(mpg ~ cyl, data = mtcars)

# 3. Gaussian GLM
glm_model <- glm(mpg ~ cyl, family = gaussian(link = "identity"), data = mtcars)

# Compare coefficients
coef_comparison <- data.frame(
  Term = names(coef(lm_model)),
  `Linear Regression` = coef(lm_model),
  `Gaussian GLM` = coef(glm_model)
)

# Display the comparison
coef_comparison %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  flextable()

# Compare ANOVA tables
anova_aov <- anova(anova_model)
anova_lm <- anova(lm_model)
anova_glm <- anova(glm_model)

# Create visualization showing the three approaches
# Use the same data and estimated means
ggplot() +
  # Plot raw data
  geom_boxplot(data = mtcars, 
              aes(x = cyl, y = mpg, group = cyl),
              alpha = 0.3, width = 0.5) +
  geom_jitter(data = mtcars, 
             aes(x = cyl, y = mpg),
             width = 0.1, alpha = 0.6) +
  # Add fitted values from each model
  geom_point(data = emmeans(lm_model, ~cyl) %>% data.frame(),
            aes(x = cyl, y = emmean), 
            color = "red", size = 3, shape = 17) +
  geom_point(data = emmeans(glm_model, ~cyl) %>% data.frame(),
            aes(x = cyl, y = emmean), 
            color = "blue", size = 3, shape = 15) +
  # Add legend for model types
  annotate("text", x = 8, y = 30, 
          label = "Red triangles: Linear Regression\nBlue squares: Gaussian GLM", 
          hjust = 1, size = 3.5) +
  labs(title = "Comparison of Models: ANOVA, Linear Regression, and Gaussian GLM",
       subtitle = "All three approaches yield identical results",
       x = "Number of Cylinders",
       y = "Miles Per Gallon") +
  theme_minimal()
```
:::

# Assumptions and Diagnostics Summary

::::: columns
::: {.column width="60%"}
-   Generalized Linear Models- assumptions depend on distribution+link
    function used:
-   **All GLMs:**
    -   Independence of observations, No outliers, No Multicolinearity
    -   Correct specification of the link function /variance structure
-   **Gaussian GLMs (including linear regression):**
    -   Normality of residuals
    -   Homogeneity of variance
-   **Poisson GLMs:**
    -   Count data (non-negative integers)
    -   Mean equals variance - overdispersed = negative binomial)
-   **Logistic GLMs:**
    -   Binary response variable
    -   Linear relationship between predictors and log odds
    -   Adequate sample size relative to number of parameters
:::

::: {.column width="40%"}
R code checks common diagnostics for logistic model:

```{r diagnostic-summary}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 6
#| paged-print: false
# Create diagnostic plots for the rodent model
par(mfrow = c(2, 2))

# 1. Residuals vs fitted
plot(fitted(rodent_model), residuals(rodent_model, type = "pearson"),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values (predicted probabilities)",
     ylab = "Pearson Residuals",
     pch = 16)
abline(h = 0, lty = 2)

# 2. Leverage
leverage <- hatvalues(rodent_model)
plot(leverage, residuals(rodent_model, type = "pearson"),
     main = "Residuals vs Leverage",
     xlab = "Leverage",
     ylab = "Pearson Residuals",
     pch = 16)
abline(h = 0, lty = 2)

# 3. Cook's distance
cook <- cooks.distance(rodent_model)
plot(cook, main = "Cook's Distance",
     ylab = "Cook's Distance",
     pch = 16)
abline(h = 4/length(cook), lty = 2, col = "red")  # Rule of thumb threshold

# 4. Observed vs Predicted probabilities
plot(predicted_probs, 
     as.numeric(fragment_data$rodent_present) - 1,
     main = "Observed vs Predicted",
     xlab = "Predicted Probability",
     ylab = "Observed (0/1)",
     pch = 16)
curve(I, from = 0, to = 1, add = TRUE, col = "red")
```
:::
:::::

# Summary and Conclusions

Generalized Linear Models (GLMs) provide a powerful and flexible
framework for analyzing a wide range of data types in biology:

1.  **Gaussian GLMs** identity link function = linear model ANOVAs

2.  **Poisson GLMs** log link function appropriate for count data,
    cautious of overdispersion.

3.  **Negative Binomial** works with overdispersed data

4.  **Logistic GLMs** with logit link function useful for binary
    responses

    -   probability of success or presence.

-   Key advantages of GLMs include:
    -   Handle various response variables beyond normal distributions
    -   Unified framework for linear modeling
    -   Flexibility in specifying the link function to match the data
        structure
    -   Interpretable parameters, though interpretation differs by model
        type

# Summary and Conclusions

When working with GLMs:

1.  Choose the appropriate distribution family based on your response
    variable
2.  Verify model assumptions through diagnostic plots
3.  Watch for overdispersion in count data
4.  Use odds ratios to interpret logistic regression results
5.  Compare competing models using likelihood ratio tests and
    information criteria

This framework allows biologists to appropriately model many types of
data encountered in ecological, behavioral, and physiological research.

# References

-Agresti, A. (1996). An Introduction to Categorical Data Analysis.
Wiley, New York. Bolger, D. T., Alberts, A. C., Sauvajot, R. M.,
Potenza, P., McCalvin, C., Tran, D., Mazzoni, S., & Soulé, M. E. (1997).
Response of rodents to habitat fragmentation in coastal southern
California. Ecological Applications, 7(2), 552-563. -Christensen, R.
(1997). Log-linear Models and Logistic Regression. Springer, New York.
-Hosmer, D. W., & Lemeshow, S. (1989). Applied Logistic Regression.
Wiley, New York. -McCullagh, P., & Nelder, J. A. (1989). Generalized
Linear Models. Chapman and Hall, London. -Polis, G. A., Hurd, S. D.,
Jackson, C. T., & Piñero, F. S. (1998). Multifactor analysis of
ecosystem patterns on islands in the Gulf of California. Ecological
Monographs, 68, 490-502.
